Towards a Verified Tableau Prover for Multi-Level-Syllogistics

### 2022-08-29 paper/paper.tex
Continue

\newcommand{\sexpands}[2]{#1 $\lefttriangle$ #2}
\newcommand{\fexpands}[2]{#1 $\lefttrianglebar$ #2}
\newcommand{\expandss}[2]{#1 $\lefttriangle^*$ #2}

@@

type_synonym 'a pset_fm = 'a pset_atom fm
\end{lstlisting}
We will sometimes drop the \lstinline!Atom! constructor to reduce clutter.
Additionally, we use the abbreviations \lstinline!s $\notinS$ t! and \lstinline!s $\neqS$ t! to denote \lstinline!&\fmNeg{\fmAtom{} (s $\inS$ t)}&! and \lstinline!&\fmNeg{\fmAtom{} (s $\eqS$ t)}&!, respectively.
Note that Isabelle's datatype package automatically defines a function \lstinline!atoms :: 'a fm $\Rightarrow$ 'a set! that allows us to obtain all set atoms in a formula.
Similarly, it defines a function \lstinline!set_pset_atom :: 'a pset_atom $\Rightarrow$ 'a set! with which we can extract all the variable names in a set atom.
We combine these functions in order to define \lstinline!vars_fm! that extracts all the variable names occuring in a set formula.
\todo[inline]{Motivation: one rule introduces new variables}
\begin{lstlisting}
definition vars_fm :: 'a pset_fm $\Rightarrow$ 'a set where
  vars_fm $\phi$ $\equiv$ $\bigcup$(set_pset_atom ` atoms $\phi$)
\end{lstlisting}
In addition to the variable names that appear in a formula, we will later also need to obtain those set terms that occur as subterms in a formula. 
We first define a function that collects the subterms of a set term.
\todo[inline]{Motivation: saturating rules don't introduce subterms}
\begin{lstlisting}
fun subterms ::
    'a pset_term $\Rightarrow$ 'a pset_term set where
  subterms $\emptyset$ = {$\emptyset$}
| subterms (Var x) = {Var x}
| subterms (s $\unionS$ t) =
    {s $\unionS$ t} $\cup$ subterms s $\cup$ subterms t
| subterms (s $\interS$ t) =
    {s $\interS$ t} $\cup$ subterms s $\cup$ subterms t
| subterms (s $\diffS$ t) =
    {s $\diffS$ t} $\cup$ subterms s $\cup$ subterms t
| subterms (Single t) =
    {Single t} $\cup$ subterms t
\end{lstlisting}
The subterms of an atom are then the subterms of those terms that make up the atom.
\begin{lstlisting}
fun subterms_atom ::
      'a pset_atom $\Rightarrow$ 'a pset_term set where
  subterms_atom (s $\inS$ t) =
    subterms s $\cup$ subterms t
| subterms_atom (s $\eqS$ t) =
    subterms s $\cup$ subterms t
\end{lstlisting}
Ultimately, we lift this to formulae by using \lstinline!atoms! again.
\begin{lstlisting}
definition subterms_fm ::
    'a pset_fm $\Rightarrow$ 'a pset_term set where
  subterms_fm $\phi$ $\equiv$ $\bigcup$(subterms_atom ` atoms $\phi$)
\end{lstlisting}
The last function that we define computes the subformulae of a formula.
\begin{lstlisting}
fun subfms :: 'a fm $\Rightarrow$ 'a fm set where
  subfms (&\fmAtom& a) = {&\fmAtom& a}
| subfms (&\fmAnd{p}{q}&) =
    {&\fmAnd{p}{q}&} $\cup$ subfms p $\cup$ subfms q
| subfms (&\fmOr{p}{q}&) =
    {&\fmOr{p}{q}&} $\cup$ subfms p $\cup$ subfms q
| subfms (&\fmNeg{q}&) = {&\fmNeg{q}&} $\cup$ subfms q
\end{lstlisting}

@@

In Isabelle/HOL, which is simply typed, this definition is not accepted, though.
A way to work around this limitiation is provided by Lawrence in an AFP entry~\cite{zfc_in_hol_afp}, which adds $\mathcal{V}$ to Isabelle/HOL by way of axiomatisation.
An article by Lawrence~\cite{zfc_in_hol} provides some further context to this entry.
The entry declares a type \lstinline!V! that is equipped with a function \lstinline!elts :: V $\Rightarrow$ V set! that maps a set to its elements.
Even though we use the typed sets provided by Isabelle/HOL here, \lstinline!elts! actually returns a class.
Only if the class is small enough, as indicated by the polymorphic predicate \lstinline!small :: 'a set $\Rightarrow$ bool!, it is indeed a set.
Inversely, this predicate characterises those sets that can be turned into a set of type \lstinline!V! using the polymorphic function \lstinline!vset :: 'a set $\Rightarrow$ V!.
The type \lstinline!V! comes with all the usual set operations such as equality ($=$), union ($\sqcup$), intersection ($\sqcap$), and difference ($-$).
The empty set coincides with the ordinal $0$ so we denote it with \lstinline!0 :: V!.

@@

definition vars_branch :: 'a branch $\Rightarrow$ 'a set 
  vars_branch b $\equiv$ $\bigcup$(vars_fm ` set b)
\end{lstlisting}
In the standard tableau calculus for propositional logic, a branch is called \textit{closed} if it contains both the negation of a formula as well as the formula itself and it is called \textit{open} if it is not closed.
We extend the notion of closedness with three additional rules; the first two are straightforward while the last one states that a branch is closed when the branch contains a membership cycle \lstinline!s $\inS$ t$_\text{1}$, t$_\text{1}$ $\inS$ t$_\text{2}$, $\ldots$, t$_\text{k}$ $\inS$ s!.
\begin{lstlisting}
inductive bclosed :: 'a branch $\Rightarrow$ bool where
  $\llbracket$ $\phi$ $\in$ set b; &\fmNeg{$\phi$}& $\in$ set b $\rrbracket$ $\Longrightarrow$ bclosed b
| (t $\inS$ $\emptyset$) $\in$ set b $\Longrightarrow$ bclosed b
| (t $\neqS$ t) $\in$ set b $\Longrightarrow$ bclosed b
| $\llbracket$ member_cycle cs; set cs $\subseteq$ set b$\rrbracket$
    $\Longrightarrow$ bclosed b

@@

\subsection{Saturating Expansion Rules}
The calculus considers two kinds of expansion rules: \textit{saturating} and \textit{fulfilling}.
Whereas fulfilling expansion rules lead to the creation of new branches in the tableau, saturating rules only extend a branch \lstinline!b! with new formulae \lstinline[breaklines=true]!b' = [$\psi_1$,$\ldots$,$\psi_n$]!, which we denote by \lstinline!&\sexpands{b'}{b}&!.
Furthermore, saturating rules are restricted such that no new subterm may be created by their application.
For example, the rule
\begin{center}
\lstinline!(s $\inS$ t$_\text{1}$) $\in$ set b $\Longrightarrow$ &\sexpands{[s $\inS$ t$_\text{1}$ $\unionS$ t$_\text{2}$]}{b}&!
\end{center}
may only be used under the condition
\begin{center}
  \lstinline!t$_\text{1}$ $\unionS$ t$_\text{2}$ $\in$ subterms_fm (last b)!.
\end{center}
This ensures that repeated application of saturating rules eventually results in a \textit{linearly saturated} branch, i.e.\ a branch where no application of saturating rules would produce new formulae.
In Isabelle, we express this as follows.
\begin{lstlisting}
definition lin_sat b $\equiv$
  $\forall$b'. &\sexpands{b'}{b}& $\longrightarrow$ set b' $\subseteq$ set b
\end{lstlisting}
\todo[inline]{Reference figure and mention missing rule + example}

\subsection{Fulfilling Expansion Rules\label{sec:fulfilling}}
Since the number of branches doubles each time we apply a fulfilling rule, it is prudent to only apply them to open, linearly saturated branches. 
To further prevent unnecessary branching, each rule has a \textit{subsumption} condition.

@@

\section{Completeness of the Calculus}

\begin{lstlisting}
definition wf_branch b $\equiv$ $\exists$$\phi$. &\expandss{b}{[$\phi$]}&
\end{lstlisting}

@@

\section{An Abstract Specification of the Decision Procedure}
Think about using the notion of coherent (or is this wfbranch?)
\begin{figure*}
  \centering
\begin{lstlisting}
function mlss_proc_branch :: 'a branch $\Rightarrow$ bool where
  $\neg$ lin_sat b $\Longrightarrow$
  mlss_proc_branch = mlss_proc_branch ((SOME b'. &\sexpands{b'}{b}& $\land$ set b $\subset$ set (b' @ b)) @ b)
|  $\llbracket$ $\neg$ sat b; bopen b; lin_sat b $\rrbracket$ $\Longrightarrow$
   mlss_proc_branch b = ($\forall$b' $\in$ (SOME bs. &\fexpands{bs}{b}&). mlss_proc_branch (b' @ b))
| $\llbracket$ lin_sat b; sat b $\rrbracket$ $\Longrightarrow$ mlss_proc_branch b = bclosed b
| $\llbracket$ lin_sat b; bclosed b $\rrbracket$ $\Longrightarrow$ mlss_proc_branch b = True

@@

\subsection{Termination}
Maybe interesting: technical lemma that allows to obtain the point at which a parameter was introduced.
\subsection{Soundness}
Interesting: The model changes because auf new parameters.
Maybe move this to the chapter on the calculus?
\subsection{Lifting the Completeness of the Calculus}
\todo[inline]{Do you even lift?}

>>>>

\newcommand{\lexpands}[2]{#1 $\lefttriangle$ #2}
\newcommand{\fexpands}[2]{#1 $\lefttrianglebar$ #2}
\newcommand{\expandss}[2]{#1 $\lefttriangle^*$ #2}

@@

type_synonym 'a pset_fm = 'a pset_atom fm
\end{lstlisting}
We will sometimes drop the \lstinline!Atom! constructor to reduce clutter.
Additionally, we use the abbreviations \lstinline!s $\notinS$ t! and \lstinline!s $\neqS$ t! to denote \lstinline!&\fmNeg{\fmAtom{} (s $\inS$ t)}&! and \lstinline!&\fmNeg{\fmAtom{} (s $\eqS$ t)}&!, respectively.
Note that Isabelle's datatype package automatically defines a function \lstinline!atoms :: 'a fm $\Rightarrow$ 'a set! that allows us to obtain all set atoms in a formula.
Similarly, it defines a function \lstinline!set_pset_atom :: 'a pset_atom $\Rightarrow$ 'a set! with which we can extract all the variable names in a set atom.
We combine these functions in order to define \lstinline!vars_fm! that extracts all the variable names occuring in a set formula.
\todo[inline]{Motivation: one rule introduces new variables}
\begin{lstlisting}
definition vars_fm :: 'a pset_fm $\Rightarrow$ 'a set where
  vars_fm $\phi$ $\equiv$ $\bigcup$(set_pset_atom ` atoms $\phi$)
\end{lstlisting}
In addition to the variable names that appear in a formula, we will later also need to obtain those set terms that occur as subterms in a formula. 
We first define a function that collects the subterms of a set term.
\todo[inline]{Motivation: linear rules don't introduce subterms}
\begin{lstlisting}
fun subterms ::
    'a pset_term $\Rightarrow$ 'a pset_term set where
  subterms $\emptyset$ = {$\emptyset$}
| subterms (Var x) = {Var x}
| subterms (s $\unionS$ t) =
    {s $\unionS$ t} $\cup$ subterms s $\cup$ subterms t
| subterms (s $\interS$ t) =
    {s $\interS$ t} $\cup$ subterms s $\cup$ subterms t
| subterms (s $\diffS$ t) =
    {s $\diffS$ t} $\cup$ subterms s $\cup$ subterms t
| subterms (Single t) =
    {Single t} $\cup$ subterms t
\end{lstlisting}
The subterms of an atom are then the subterms of those terms that make up the atom.
\begin{lstlisting}
fun subterms_atom ::
      'a pset_atom $\Rightarrow$ 'a pset_term set where
  subterms_atom (s $\inS$ t) =
    subterms s $\cup$ subterms t
| subterms_atom (s $\eqS$ t) =
    subterms s $\cup$ subterms t
\end{lstlisting}
Ultimately, we lift this to formulae by using \lstinline!atoms! again.
\begin{lstlisting}
definition subterms_fm ::
    'a pset_fm $\Rightarrow$ 'a pset_term set where
  subterms_fm $\phi$ $\equiv$ $\bigcup$(subterms_atom ` atoms $\phi$)
\end{lstlisting}
The last function that we define computes the subformulae of a formula.
\begin{lstlisting}
fun subfms :: 'a fm $\Rightarrow$ 'a fm set where
  subfms (&\fmAtom& a) = {&\fmAtom& a}
| subfms (&\fmAnd{p}{q}&) =
    {&\fmAnd{p}{q}&} $\cup$ subfms p $\cup$ subfms q
| subfms (&\fmOr{p}{q}&) =
    {&\fmOr{p}{q}&} $\cup$ subfms p $\cup$ subfms q
| subfms (&\fmNeg{q}&) = {&\fmNeg{q}&} $\cup$ subfms q
\end{lstlisting}

@@

In Isabelle/HOL, which is simply typed, this definition is not accepted, though.
A way to work around this limitiation is provided by Lawrence in an AFP entry~\cite{zfc_in_hol_afp}, which adds $\mathcal{V}$ to Isabelle/HOL by way of axiomatisation.
An article by Lawrence~\cite{zfc_in_hol} provides some further context to this entry.
The entry declares a type \lstinline!V! that is equipped with a function \lstinline!elts :: V $\Rightarrow$ V set! that maps a set to its elements.
Even though we use the typed sets provided by Isabelle/HOL here, \lstinline!elts! actually returns a class.
Only if the class is small enough, as indicated by the polymorphic predicate \lstinline!small :: 'a set $\Rightarrow$ bool!, it is indeed a set.
Inversely, this predicate characterises those sets that can be turned into a set of type \lstinline!V! using the polymorphic function \lstinline!vset :: 'a set $\Rightarrow$ V!.
The type \lstinline!V! comes with all the usual set operations such as equality ($=$), union ($\sqcup$), intersection ($\sqcap$), and difference ($-$).
The empty set coincides with the ordinal $0$ so it is denoted by \lstinline!0 :: V!.

@@

definition vars_branch :: 'a branch $\Rightarrow$ 'a set 
  vars_branch b $\equiv$ $\bigcup$(vars_fm ` set b)
\end{lstlisting}
In the standard tableau calculus for propositional logic, a branch is called \textit{closed} if it contains both the negation of a formula and the formula itself;
conversely, it is called \textit{open} if it is not closed.
For MLSS, we extend the notion of closedness with three additional rules; the first two are straightforward while the last one states that a branch is closed when the branch contains a membership cycle \lstinline!s $\inS$ t$_\text{1}$, t$_\text{1}$ $\inS$ t$_\text{2}$, $\ldots$, t$_\text{k}$ $\inS$ s!.
\begin{lstlisting}
inductive bclosed :: 'a branch $\Rightarrow$ bool where
  $\llbracket$ $\phi$ $\in$ set b; &\fmNeg{$\phi$}& $\in$ set b $\rrbracket$ $\Longrightarrow$ bclosed b
| (t $\inS$ $\emptyset$) $\in$ set b $\Longrightarrow$ bclosed b
| (t $\neqS$ t) $\in$ set b $\Longrightarrow$ bclosed b
| $\llbracket$ member_cycle cs; set cs $\subseteq$ set b$\rrbracket$
    $\Longrightarrow$ bclosed b

@@

\subsection{Linear Expansion Rules}
The calculus considers two kinds of branch expansion rules: \textit{linear} and \textit{branching} rules.
As the name suggests, branching rules lead to the creation of new branches in the tableau while linear rules only extend a branch \lstinline!b! with new formulae \lstinline[breaklines=true]!b' = [$\psi_1$,$\ldots$,$\psi_n$]!, which we denote by \lstinline!&\lexpands{b'}{b}&!.
Furthermore, linear rules are restricted such that no new subterm may be created by their application:
for example, the rule
\begin{center}
\lstinline!(s $\inS$ t$_\text{1}$) $\in$ set b $\Longrightarrow$ &\lexpands{[s $\inS$ t$_\text{1}$ $\unionS$ t$_\text{2}$]}{b}&!
\end{center}
may only be used under the condition
\begin{center}
  \lstinline!t$_\text{1}$ $\unionS$ t$_\text{2}$ $\in$ subterms_fm (last b)!.
\end{center}
\todo[inline]{Explain this restriction in more detail.}
This ensures that repeated application of linear rules eventually results in a \textit{linearly saturated} branch, i.e.\ a branch where no application of linear rules would produce new formulae.
In Isabelle, we express this as follows.
\begin{lstlisting}
definition lin_sat b $\equiv$
  $\forall$b'. &\lexpands{b'}{b}& $\longrightarrow$ set b' $\subseteq$ set b
\end{lstlisting}
\todo[inline]{Reference figure and mention missing rule + example}

\subsection{Branching Rules\label{sec:branching}}
Since the number of branches increases when applying branching rules, it is prudent to only apply them to open, linearly saturated branches.
To further prevent unnecessary branching, each rule has a \textit{subsumption} condition. 
Note that even branching rules are restricted such that they never create new subterms, with exception of the last rule which adds a fresh variable.
\todo[inline]{Reference figure}

@@

\section{Completeness of the Calculus}
\begin{itemize}
  \item Basic definitions and graph
  \item Graph is basically a syntactic model.
\end{itemize}
\begin{lstlisting}
definition wf_branch b $\equiv$ $\exists$$\phi$. &\expandss{b}{[$\phi$]}&
\end{lstlisting}

@@

\section{An Abstract Specification of the Decision Procedure}
\todo[inline]{Mention coherence lemma?}
\begin{figure*}
  \centering
\begin{lstlisting}
function mlss_proc_branch :: 'a branch $\Rightarrow$ bool where
  $\neg$ lin_sat b $\Longrightarrow$
  mlss_proc_branch = mlss_proc_branch ((SOME b'. &\lexpands{b'}{b}& $\land$ set b $\subset$ set (b' @ b)) @ b)
|  $\llbracket$ $\neg$ sat b; bopen b; lin_sat b $\rrbracket$ $\Longrightarrow$
   mlss_proc_branch b = ($\forall$b' $\in$ (SOME bs. &\fexpands{bs}{b}&). mlss_proc_branch (b' @ b))
| $\llbracket$ lin_sat b; sat b $\rrbracket$ $\Longrightarrow$ mlss_proc_branch b = bclosed b
| $\llbracket$ lin_sat b; bclosed b $\rrbracket$ $\Longrightarrow$ mlss_proc_branch b = True

@@

\subsection{Termination}
Maybe interesting: technical lemma that allows to obtain the point at which a parameter was introduced.
\subsection{Soundness}
Interesting: The model changes because auf new parameters.
Maybe move this to the chapter on the calculus?
\todo[inline]{Show \lstinline!memberCycle! case? This requires \lstinline!member_seq!.}
\subsection{Lifting the Completeness of the Calculus}
\todo[inline]{Do you even lift?}
